---
title: "Mobility and Transportation model"
author: Roletto Enrico
date: 18/May/2018
output: html_notebook
---

# Invidivual Project for the course in Smart City and IoT @MADAS 2017/2018

For the project we need these libraries
```{r First_things, message=FALSE, warning=FALSE, paged.print=FALSE}
rm(list=ls())
gc()

library(data.table)
library(igraph)
library(stringr)
library(readxl)

```

Loading the data from the excel spreedsheet
```{r Dataimp, message=FALSE, warning=FALSE}
datamap <- as.data.table(read_excel("Transportmodeling.xlsx", 
                           sheet = "DATAMAP"))
network <- as.data.table(read_excel("Transportmodeling.xlsx", 
                      sheet = "NETWORK"))
coord <- as.data.table(read_excel("Transportmodeling.xlsx", 
                      sheet = "COORD"))
speedFlow <- as.data.table(read_excel("Transportmodeling.xlsx", 
                    sheet = "SPEED-FLOW"))

matOD <- as.data.table(read_excel("Transportmodeling.xlsx", 
                                      sheet = "MAT_OD"))
```

Let's have a look at the speed flow table. In this table we have the percetange reduction in speed given a capacity index and a load level (numver of vehicles on the roads over the capacity). We will reshape these data in order to use them easely later:

- we change the shape from wide to long
- we create two columns representing the inferior and superior limit for each row.
  

```{r speed_flow}
head(speedFlow)

speedFlow[, SPEED_0:=0]

speedFlow<- melt(speedFlow, id=1, measure.vars = c(2:7), 
                 variable.name = 'load_left', value.name = 'speed_red')[,
            load_left := as.numeric(str_extract(load_left, "([0-9])"))/10] 

setorder(speedFlow, CAP_INDEX, load_left)

speedFlow[, load_right:=shift(load_left,1, type = 'lead'), by=CAP_INDEX]
speedFlow[is.na(speedFlow)]<- 10

head(speedFlow, 3)
tail(speedFlow, 3)

```

Let's have now a look at the origin destination table. 

- We remove the duplicated rows, adding up the vehicles among duplicates
- I create 3 new columns each one with anumber of vehicles according to the i rulncremental 30% - 20% of the total.


```{r OD_mat}
sum(duplicated(matOD[,.(ZONE_O, ZONE_D)]))
matOD <- matOD[ZONE_O!=ZONE_D, .(VEH_H=sum(VEH_H)), by=c('ZONE_O', 'ZONE_D')]

matOD[, c('VEH_50', 'VEH_30', 'VEH_20'):= .(ceiling(VEH_H*0.5), round(VEH_H*0.3), VEH_H - round(VEH_H*0.3) - ceiling(VEH_H*0.5))]

names(network) <- gsub(x = names(network),
                        pattern = "-",
                        replacement = "_")
```

